A Geometry for Network Visualization in \MakeLowercase{\texttt{ggplot2}}

### 2015-04-02 CC/writeup.Rnw
add data, installation instructions, R files, and examples :wq

In order to construct the \texttt{geom} in \texttt{ggplot2}, we needed two \texttt{R} script files: a `stat' file and a `geom' file. The stat file performs all the necessary calculations on the two input data frames, and the geom file performs the drawing of the network.
  \subsection{Creating \texttt{stat\_net}}
  In order to plot the network with only the edge connections and vertex names, we relied on the \texttt{sna} and \texttt{network} packages \citep{sna, network.jss}. First, we used the edges data frame to construct a network object with the \texttt{as.network()} function, then constructed an adjacency matrix from that network using the \texttt{as.matrix.network.adjacency()} function.  This adjacency matrix is then passed to the chosen \texttt{gplot.layout.*()} function from the \texttt{sna} package.  \emph{When the layout argument is changed in the \texttt{geom\_net()} function, the corresponding \texttt{gplot.layout.*()} function in \texttt{sna} is called with a \texttt{do.call()} statement.} This layout function produces a matrix of coordinates of the vertices, which we then transform into a data frame containing the coordinates of the edges using the \texttt{as.matrix.network.edgelist()} function in \texttt{network}.  We are grateful to Moritz Marbach for his \texttt{gplot()} function which we used when creating this process \citep{marbach}.  At the end of the previous routine, we have a data frame with \texttt{x, y, xend,} and \texttt{yend} columns describing the edges and another data frame with \texttt{x,y} columns describing the vertices.  These two data frames, along with the many \texttt{aes()} and other argument options get passed on to the actual network geometry for drawing.
   \subsection{\texttt{stat\_net} Usage}
<<usage_stat_net, eval=FALSE>>=
stat_net(mapping = NULL, data = NULL, vertices = NULL, geom = "net", position = "identity", ...)
#aes(colour = "black", size =3, layout="kamadakawai",layout.par = NA)
@
  The usage of the \texttt{stat\_net} arguments is as follows:
  \begin{itemize}
  \item The \texttt{mapping} argument is the aesthetic mapping. Currently, there is no reason to ever change it from \texttt{NULL}.
  \item The \texttt{data} argument is the edges data frame.  This is sent to \texttt{stat\_net} through the \texttt{ggplot()} function.
  \item The \texttt{vertices} argument is the vertex data frame. This is sent to \texttt{stat\_net} through the \texttt{geom\_net} function.
  \item The \texttt{geom} argument is the geometric object used to plot the data. It is by default set to \texttt{net}. 
  \item The \texttt{position} argument is set to \texttt{identity} because we wish to plot the network as-is, with no adjustments.
  \item na.rm
  \end{itemize}
  \texttt{stat\_net} also accepts the following aesthetic values, none of which are required:
  \begin{itemize}
  \item The \texttt{layout} aesthetic is the layout algorithm to use to place the vertices of the network.  The default is the Kamada-Kawai algorithm, and there are other options listed in \ref{table:layouts}.  
  \item The \texttt{layout.par} aesthetic is the layout arguments to pass to the layout function.  It is set to \texttt{NA} by default, in which case the algorithm uses its set defaults in the \texttt{sna} package. This aesthetic takes a list of parameters according to the documentation by \citet{sna}. 
  \end{itemize}

@@

   \subsection{\texttt{geom\_net} Usage}
<<usage_geom_net, eval=FALSE>>=
geom_net(mapping = NULL, data = NULL, vertices = NULL, vlabel = FALSE, directed = FALSE, stat = "net", ...)
#aes(ecolour = "grey20", vcolour = "black", vfill="black", fill = NA, esize = 1, vsize = 2, elinetype = 1, weight = NA, valpha = 1, vshape = 16, ealpha=1)
@
  The usage of the \texttt{geom\_net} arguments is as follows:
  \begin{itemize}
  \item The \texttt{mapping} argument is the aesthetic mapping. It is constructed with \texttt{aes()} according to the aesthetic values that can be passed to \texttt{geom\_net}. 
  \item The \texttt{data} argument is the edges data frame.  There is usually no need to use it in \texttt{geom\_net()} because it will have already been provided in the \texttt{ggplot()} function.
  \item The \texttt{vertices} argument is the vertex data frame. The vertex data frame should be provided to this argument in \texttt{geom\_net}. This data frame is required to have a column named \texttt{label} providing the labels of all vertices in the network.
  \item The \texttt{vlabel} argument is a logical value which plots the label layer on top of the edge and vertex layers when it is set to \texttt{TRUE}. 
  \item The \texttt{directed} argument is a logical value which adds arrows to the edge layer pointing to the \texttt{to\_id} vertex when it is set to \texttt{TRUE}.  
  \item The \texttt{stat} argument is the statistical transformation used on the data for the network layer. It is always set to \texttt{net}. 
  \end{itemize}
  \texttt{geom\_net} also accepts the following aesthetic values, and the ones marked with an asterisk are required:
  \begin{itemize}
  \item The \texttt{from\_id}* and \texttt{to\_id}* aesthetics each take one column which, when combined, provide the edges in a network.  If the network is undirected, then their order does not matter. If the network is directed, then the \emph{from} vertex names should all be in the \texttt{from\_id} column, and their corresponding \emph{to} vertex names should all be in the \texttt{to\_id} column. 
  \item The \texttt{ecolour} aesthetic changes the color of the edges. Its default is \texttt{grey20}
  \item The \texttt{esize} aesthetic changes the size (thickness) of the edges. Its default is 1. 
  \item The \texttt{elinetype} aesthetic changes the linetype of the edges. Its default is 1. 
  \item The \texttt{ealpha} aesthetic changes the transparency of the edges. Its default is 1. 
  \item The \texttt{vcolour} aesthetic changes the color of the vertices. Its default is \texttt{black}. 
  \item The \texttt{vfill} aesthetic changes the ouline color of the vertices. Its default is \texttt{black}. 
  \item The \texttt{vsize} aesthetic changes the size (diameter) of the vertices. Its default is 2.
  \item The \texttt{vshape} aesthetic changes the shape of the vertices. Its default is 16 (solid circular points). 
  \item The \texttt{valpha} aesthetic changes the transparency of the vertices. Its default is 1. 
  \end{itemize}

>>>>

In order to construct the \texttt{geom} in \texttt{ggplot2}, we needed two \texttt{R} script files: a `stat' file and a `geom' file. The stat file performs all the necessary calculations on the two input data frames, and the geom file performs the drawing of the network.
  \subsection{Creating \texttt{stat\_net}}
  In order to plot the network with only the edge connections and vertex names, we relied on the \texttt{sna} and \texttt{network} packages \citep{sna, network.jss}. First, we used the edges data frame to construct a network object with the \texttt{as.network()} function, then constructed an adjacency matrix from that network using the \texttt{as.matrix.network.adjacency()} function.  This adjacency matrix is then passed to the chosen \texttt{gplot.layout.*()} function from the \texttt{sna} package.  \emph{When the layout argument is changed in the \texttt{geom\_net()} function, the corresponding \texttt{gplot.layout.*()} function in \texttt{sna} is called with a \texttt{do.call()} statement.} This layout function produces a matrix of coordinates of the vertices, which we then transform into a data frame containing the coordinates of the edges using the \texttt{as.matrix.network.edgelist()} function in \texttt{network}.  We are grateful to Moritz Marbach for his \texttt{gplot()} function which we used when creating this process \citep{marbach}.  At the end of the previous routine, we have a data frame with \texttt{x, y, xend,} and \texttt{yend} columns describing the edges and another data frame with \texttt{x,y} columns describing the vertices.  These two data frames, along with the many \texttt{aes()} and other argument options get passed on to the actual network geometry for drawing.
   \subsection{\texttt{stat\_net} Usage}
<<usage_stat_net, eval=FALSE>>=
stat_net(mapping = NULL, data = NULL, vertices = NULL, geom = "net", position = "identity", ...)
@
  The usage of the \texttt{stat\_net} arguments is as follows:
  \begin{itemize}
  \item The \texttt{mapping} argument is the aesthetic mapping. Currently, there is no reason to ever change it from \texttt{NULL}.
  \item The \texttt{data} argument is the edges data frame.  This is sent to \texttt{stat\_net} through the \texttt{ggplot()} function.
  \item The \texttt{vertices} argument is the vertex data frame. This is sent to \texttt{stat\_net} through the \texttt{geom\_net} function.
  \item The \texttt{geom} argument is the geometric object used to plot the data. It is by default set to \texttt{net}. 
  \item The \texttt{position} argument is set to \texttt{identity} because we wish to plot the network as-is, with no adjustments.
  \item na.rm
  \end{itemize}
  \texttt{stat\_net} also accepts the following aesthetic values, none of which are required:
  \begin{itemize}
  \item The \texttt{layout} aesthetic is the layout algorithm to use to place the vertices of the network.  The default is the Kamada-Kawai algorithm, and there are other options listed in \ref{table:layouts}.  
  \item The \texttt{layout.par} aesthetic is the layout arguments to pass to the layout function.  It is set to \texttt{NA} by default, in which case the algorithm uses its set defaults in the \texttt{sna} package. This aesthetic takes a list of parameters according to the documentation by \citet{sna}. 
  \end{itemize}

@@

   \subsection{\texttt{geom\_net} Usage}
<<usage_geom_net, eval=FALSE>>=
geom_net(mapping = NULL, data = NULL, vertices = NULL, vlabel = FALSE, directed = FALSE, stat = "net", ...)
@
  The usage of the \texttt{geom\_net} arguments is as follows:
  \begin{itemize}
  \item The \texttt{mapping} argument is the aesthetic mapping. It is constructed with \texttt{aes()} according to the aesthetic values that can be passed to \texttt{geom\_net}. 
  \item The \texttt{data} argument is the edges data frame.  There is usually no need to use it in \texttt{geom\_net()} because it will have already been provided in the \texttt{ggplot()} function.
  \item The \texttt{vertices} argument is the vertex data frame. The vertex data frame should be provided to this argument in \texttt{geom\_net}. This data frame is required to have a column named \texttt{label} providing the labels of all vertices in the network.
  \item The \texttt{vlabel} argument is a logical value which plots the label layer on top of the edge and vertex layers when it is set to \texttt{TRUE}. 
  \item The \texttt{directed} argument is a logical value which adds arrows to the edge layer pointing to the \texttt{to\_id} vertex when it is set to \texttt{TRUE}.  
  \item The \texttt{stat} argument is the statistical transformation used on the data for the network layer. It is always set to \texttt{net}. 
  \end{itemize}
  \texttt{geom\_net} also accepts the following aesthetic values, and the ones marked with an asterisk are required:
  \begin{itemize}
  \item The \texttt{from\_id}* and \texttt{to\_id}* aesthetics each take one column which, when combined, provide the edges in a network.  If the network is undirected, then their order does not matter. If the network is directed, then the \emph{from} vertex names should all be in the \texttt{from\_id} column, and their corresponding \emph{to} vertex names should all be in the \texttt{to\_id} column. 
  \item The \texttt{ecolour} aesthetic changes the color of the edges. Its default is \texttt{grey20}
  \item The \texttt{esize} aesthetic changes the size (thickness) of the edges. Its default is 1. 
  \item The \texttt{elinetype} aesthetic changes the linetype of the edges. Its default is 1. 
  \item The \texttt{ealpha} aesthetic changes the transparency of the edges. Its default is 1. 
  \item The \texttt{vcolour} aesthetic changes the color of the vertices. Its default is \texttt{black}. 
  \item The \texttt{vfill} aesthetic changes the ouline color of the vertices. Its default is \texttt{black}. 
  \item The \texttt{vsize} aesthetic changes the size (diameter) of the vertices. Its default is 2.
  \item The \texttt{vshape} aesthetic changes the shape of the vertices. Its default is 16 (solid circular points). 
  \item The \texttt{valpha} aesthetic changes the transparency of the vertices. Its default is 1. 
  \end{itemize}
